# An Analysis of the Effect of Manual vs Automatic Transmissions on Gas Mileage for a Sample of Cars Build in 1973 and 1974
## Todd Flanders
### `r as.character(format(Sys.Date(), format="%B %d, %Y"))`
========================================================


```{r globalVariables, echo=FALSE}
mtcars$amChar = sapply(mtcars$am, function(x) {if(x==0) "Automatic" else "Manual"})
manual <- mtcars[mtcars$am==1,]
automatic <- mtcars[mtcars$am==0,]
meanWeight <- mean(mtcars$wt)
meanMpgManual <- mean(manual$mpg)
meanMpgAutomatic <- mean(automatic$mpg)
meanMpg <- mean(mtcars$mpg)
```


```{r echo=FALSE}
fitRow <- function(predictor) {
  meanX <- mean(mtcars[predictor][,1])
  fitAll <- lm(mtcars$mpg ~ mtcars[predictor][,1])
  fitManual <- lm(manual$mpg ~ manual[predictor][,1])
  fitAutomatic <- lm(automatic$mpg ~ automatic[predictor][,1])
  icTotal <- coef(fitAll)[1]
  beta1Total <- coef(fitAll)[2]
  icManual <- coef(fitManual)[1]
  beta1Manual <- coef(fitManual)[2]
  icAutomatic <- coef(fitAutomatic)[1]
  beta1Automatic <- coef(fitAutomatic)[2]
  mpgAtMeanManual <- icManual + (meanX * beta1Manual)
  mpgAtMeanAutomatic <- icAutomatic + (meanX * beta1Automatic)

  r <- data.frame(
    meanX,
    mpgAtMeanManual,
    mpgAtMeanAutomatic,
    icTotal,
    beta1Total,
    icManual,
    beta1Manual,
    icAutomatic,
    beta1Automatic,
    summary(fitAll)$r.squared,
    cor(fitAll$model[1], fitAll$model[2])
  )
  
  names(r) <- c("meanX", 
                "mpg@meanManual", "mpg@meanAuto",
                "icTotal", "beta1Total", 
                "icManual", "beta1Manual", 
                "icAuto", "beta1Auto", 
                "r^2", "cor")
  row.names(r) <- predictor
  
  r
}


modelRow <- function(col) {
  names(col)
}

candidateModels <- rbind(
  fitRow("cyl"),
  fitRow("disp"),
  fitRow("hp"),
  fitRow("drat"),
  fitRow("wt"),
  fitRow("qsec"),
  fitRow("vs"),
  fitRow("gear"),
  fitRow("carb")
)

bestRSquared <- candidateModels[order(-candidateModels$r^2),] 
library(ggplot2)
```

## Executive Summary
We wish to answer the following questions:
- Were manual transmissions or automatic transmissions better for gas mileage in 1974?
- What is difference measured in miles per gallon?
- What confidence do we have that our calulations represent the entire population of vehicles manufatured in 1973 and 1974?

This report analyzes data from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).  We have a small sample of `r nrow(mtcars)` cars.  `r nrow(automatic)` have automatic transmissions, and `r nrow(manual)` have manual transmissions.

## Conclusion
Although there is great desire to answer the gas milage question with certainty, the data does not lead to an indubitable conclusion.  Although correlations are strong and P-values are significant, the input data does not give confidence that the resulting model represents the population of cars manufactured in 1973 and 1974.  We would like to have multiple vehicles of similar make and model with and without automatic transmissions.  Instead, we have a very small sample which represents a wide variety of passenger, sport, luxury, and muscle cars.

## Analysis

### Marginal Values
The average mileage in the sample data is `r round(meanMpg,0)` mpg, `r round(meanMpgManual,0)` for vehicles with manual transmissions, and `r round(meanMpgAutomatic,0)` for vehicles with automatic transmissions.  If there are no counfounding effects of other variables, we expect a vehicle with a manual transmission have `r round(meanMpgManual - meanMpgAutomatic,0)` more miles per gallon than a vehicle with an automatic transmission.

### Potential Confounders
Here is a table showing the analysis of nine aspects of design and their impact on mpg for cars with automatic or manual transmissions.  The data is sorted by best model fit to worse model fit according to the r^2 statistic:

```{r results='asis', echo=FALSE}
library(xtable)
print(xtable(bestRSquared, digits=4), type="html", format.args=list(big.mark = ",", decimal.mark = "."))
```

The top three variables listed seem to make sense.  We should expect heavy cars with big engines to get worse milage than light cars with small engines.  Let's look at each variable individually.

### wt -- weight in thousands of pounds

It's no surprise that weight has the biggest impact on fuel efficiency.  More than 75% of the variance of the data is explained by weight.  

The average vehicle weight is `r round(meanWeight * 1000,0)` pounds.  This is represented by the vertical green line.  The horizontal blue line represents the marginal mileage of vehicles with manual transmissions, and the horizontal pink link represents the marginal mileage for vehicles with automatic transmissions.  The grey diagonal line represents the expected mpg for a car at a given weight, unadjusted for automatic vs manual transmission.  The pink line is the expected mpg for a car with an automatic transmission at a given weight.  The blue line is the expected mpg for a car with a manual transmission at a given weight.

```{r echo=FALSE, fig.width=7, fig.height=4}
fit <- lm(mpg ~ wt + am + wt*am, data=mtcars)
fitUnadjustedForTransType <- lm(mpg ~ wt, data=mtcars)
plot(mtcars$wt, mtcars$mpg, type = "n", frame = FALSE)
abline(fitUnadjustedForTransType, lwd = 2, col="gray")

# the marginal relationship is significant
abline(h = meanMpgManual, lwd = 3, col = "lightblue")
abline(h = meanMpgAutomatic, lwd = 3, col = "salmon")
abline(v = meanWeight, lwd=3, col = "seagreen")
abline(h = meanMpg, lwd = 3, col = "seagreen")

ica = fit$coef[1] # expected mpg of a zero wt car with auto trans
slopea = fit$coef[2]
icm = ica + fit$coef[3]
slopem = slopea + fit$coef[4]

abline(ica, slopea, lwd = 3, col="lightblue")
abline(icm, slopem, lwd = 3, col="salmon")

points(mtcars$wt[mtcars$am==0], mtcars$mpg[mtcars$am==0], pch = 21, col = "black", bg = "lightblue", cex = 2)
points(mtcars$wt[mtcars$am==1], mtcars$mpg[mtcars$am==1], pch = 21, col = "black", bg = "salmon", cex = 2)
```

We see an immediate problem in the data.  We do not have a good representation of heavy cars with automatic transmissions, nor light cars with manual transmissions.  The data is clearly biased and unreliable.

The remainder of this report will attempt to clean up the data in order to get a better representation of the effect of weight on mileage.  See appendix 1 for analysis on other variables.

The following residual plot shows that there is no obvious pattern for residuals based on the model that predicts mpg based on vehicle weight.  Perhaps residuals for all three models (automatic, manual, and all cars)

The following residual plot determines there are no interesting patterns in the residuals.  We will therefore not look for interacting variables to tighten up our parsimonious model of simple linear regression using weight and transmission type.

```{r residualPlotAll, echo=FALSE, fig.width=7, fig.height=4}
plot(fit, which=1)
```

Our three outliers are the Merc 240D, the Fiat 128 and the Toyota Corolla.  See appendix 2 for an adjusted model eliminating these outliers.  The model still has the same problem as the original.  We do not have a good representation of vehicles with different transmission types and varying weights in the original data.

## Appendix 1 -- Alternative Confounders

We selected weight as the best determination of expected miles per gallon.  Here is the analysis of variables that we did not use.

### Figure 1  -- number of cylinders

```{r echo=FALSE, fig.width=7, fig.height=4}
fit <- lm(mpg~wt, data=mtcars)
p <- ggplot(data=mtcars, aes(x=cyl, y=mpg))
p + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ amChar)
```

### Figure 2 -- engine displacement in cubic inches

```{r echo=FALSE, fig.width=7, fig.height=4}
fit <- lm(mpg~wt, data=mtcars)
p <- ggplot(data=mtcars, aes(x=disp, y=mpg))
p + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ amChar)
```

### Figure 3 -- Gross horsepower

```{r echo=FALSE, fig.width=7, fig.height=4}
fit <- lm(mpg~wt, data=mtcars)
p <- ggplot(data=mtcars, aes(x=hp, y=mpg))
p + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ amChar)
```

### Figure 4 -- Rear axle ratio

```{r echo=FALSE, fig.width=7, fig.height=4}
p <- ggplot(data=mtcars, aes(x=drat, y=mpg))
p + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ amChar)
```

### Figure 5 -- Engine configuration (V vs Straight)

```{r echo=FALSE, fig.width=7, fig.height=4}
p <- ggplot(data=mtcars, aes(x=vs, y=mpg))
p + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ amChar)
```

### Figure 6 -- Number of carburators

```{r echo=FALSE, fig.width=7, fig.height=4}
p <- ggplot(data=mtcars, aes(x=carb, y=mpg))
p + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ amChar)
```

### Figure 7 -- Number of forward gears

```{r echo=FALSE, fig.width=7, fig.height=4}
p <- ggplot(data=mtcars, aes(x=gear, y=mpg))
p + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ amChar)
```

### Figure 8 -- 1/4 mile time

```{r echo=FALSE, fig.width=7, fig.height=4}
p <- ggplot(data=mtcars, aes(x=qsec, y=mpg))
p + geom_point() + geom_smooth(method = "lm") + facet_grid(. ~ amChar)
```

## Appendix 2 -- Adjusted model eliminating the outliers Merc 240D, the Fiat 128 and the Toyota Corolla.

```{r echo=FALSE, fig.width=7, fig.height=4}
mtc <- mtcars[-c(8,18,20),]
fit <- lm(mpg ~ wt + am + wt*am, data=mtc)
fitUnadjustedForTransType <- lm(mpg ~ wt, data=mtc)
plot(mtc$wt, mtc$mpg, type = "n", frame = FALSE)
abline(fitUnadjustedForTransType, lwd = 2, col="gray")

abline(h = mean(mtc[mtc$am==0,]$mpg), lwd = 3, col = "lightblue")
abline(h = mean(mtc[mtc$am==1,]$mpg), lwd = 3, col = "salmon")
abline(v = mean(mtc$wt), lwd=3, col = "seagreen")
abline(h = mean(mtc$mpg), lwd = 3, col = "seagreen")

ica = fit$coef[1] # expected mpg of a zero wt car with auto trans
slopea = fit$coef[2]
icm = ica + fit$coef[3]
slopem = slopea + fit$coef[4]

abline(ica, slopea, lwd = 3, col="lightblue")
abline(icm, slopem, lwd = 3, col="salmon")

points(mtc$wt[mtc$am==0], mtc$mpg[mtc$am==0], pch = 21, col = "black", bg = "lightblue", cex = 2)
points(mtc$wt[mtc$am==1], mtc$mpg[mtc$am==1], pch = 21, col = "black", bg = "salmon", cex = 2)
```


```{r commonHtml, child='commonHtml.Rmd'}
```


